---
title: "2021 Olympics"
author: "Esben Lykke"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme:
      bootswatch: flatly
    keep_md: TRUE
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  cache = TRUE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 10,
  dpi = 600
)

library(pacman)

p_load(
  tidyverse,
  here,
  janitor,
  readxl,
  hrbrthemes,
  ggthemes,
  DT,
  scales,
  ggtext
)
```


### Import data
It will be a challenge to make use of all the different datasets throughout the different visualizations.
```{r import data}
medals <- read_excel(here("data", "medals.xlsx")) %>%
  clean_names() %>%
  mutate(
    team_noc = factor(team_noc),
    across(where(is.character), as.numeric)
  )

athletes <-
  read_excel(here("data", "athletes.xlsx")) %>%
  clean_names()

coaches <-
  read_excel(here("data", "coaches.xlsx")) %>%
  clean_names()

gender <-
  read_excel(here("data", "entriesgender.xlsx")) %>%
  clean_names()

teams <-
  read_excel(here("data", "teams.xlsx")) %>%
  clean_names()

pop_country <-
  read_csv(here("data", "population_by_country_2020.csv")) %>%
  clean_names()

athlete_events <-
  read_csv(here("data", "athlete_events.csv")) %>%
  clean_names()

noc_regions <-
  read_csv(here("data", "noc_regions.csv")) %>%
  clean_names()

tokyoo_2021 <-
  read_csv(here("data", "Tokyo 2021 dataset.csv")) %>%
  clean_names()
```

### Detailed overview in table
Nothing special here, just all medals information presented in a neat table.
```{r table}
medals %>%
  datatable() %>%
  formatStyle("gold", backgroundColor = "gold") %>%
  formatStyle("silver", backgroundColor = "silver") %>%
  formatStyle("bronze", backgroundColor = "orange")
```

### Total medals per country
Lets start with a common way to plot a continous variable against a categorical variable, the bar plot. 
I will try to make an emphasis on Denmark throuhgout the plots, just because. :)
```{r total medals plot}
medals_30 <-
  medals %>%
  mutate(
    team_noc = fct_reorder(team_noc, total)
  ) %>%
  slice_max(team_noc, n = 30)

medals_30 %>%
  ggplot(aes(team_noc, total)) +
  geom_segment(aes(x = team_noc, xend = team_noc, y = 0, yend = total),
    color = if_else(medals_30$team_noc == "Denmark", "orange", "grey70"),
    size = 1.2, color = "grey70"
  ) +
  geom_point(aes(team_noc, total),
    color = if_else(medals_30$team_noc == "Denmark", "orange", "grey70"),
    size = if_else(medals_30$team_noc == "Denmark", 7, 3)
  ) +
  geom_text(aes(team_noc, total, label = if_else(team_noc == "Denmark", total, NULL)), color = "white") +
  coord_flip() +
  labs(
    y = NULL,
    x = NULL,
    title = "Toal olympic medals by country: Tokyo 2021",
    subtitle = "The medal race is topped by the usual suspects, USA and China"
  ) +
  annotate("text", x = 8, y = 40, label = "With 11 medals total,\nDenmark succeeded their goal of 8-10 medals", fontface = "bold", color = "grey25") +
  annotate(
    geom = "curve", x = 9, y = 30, xend = 8, yend = 13,
    curvature = .2, arrow = arrow(length = unit(2, "mm")), color = "grey25"
  ) +
  theme_fivethirtyeight()
```

### Total medals per per capita
This shit got pretty complicated. I'm sure, there is a much easier and faster way to do this. Nevertheless, I'm pretty happy with the result so far.
```{r circular plot}
country_rank <-
  population %>%
  filter(year == 2013) %>%
  left_join(medals %>%
    rename(country = team_noc) %>%
    mutate(
      country = str_replace_all(country, c(
        "ROC" = "Russian Federation",
        "Great Britain" = "United Kingdom of Great Britain and Northern Ireland",
        "People's Republic of China" = "China"
      ))
    )) %>%
  drop_na() %>%
  mutate(
    medals_per_capita = (total / population) * 1e8,
    rank_mpc = dense_rank(desc(medals_per_capita)),
    country = str_replace_all(country, c(
      "Russian Federation" = "Russia",
      "United Kingdom of Great Britain and Northern Ireland" = "UK",
      "United States of America" = "USA",
      "San Marino" = "S. Marino"
    ))
  ) %>%
  mutate(
    name_rank = str_c(country, rank_mpc, sep = " in "),
    name_rank = case_when(
      str_detect(name_rank, c("22", "in 2")) ~ glue::glue("{name_rank}nd"),
      str_detect(name_rank, "Grenada") ~ glue::glue("{name_rank}rd"),
      str_detect(name_rank, "Marino") ~ glue::glue("{name_rank}st"),
      TRUE ~ glue::glue("{name_rank}th")
    )
  )

label_dk <-
  country_rank %>%
  filter(country == "Denmark") %>%
  mutate(
    hjust = medals_per_capita + exp(10),
    angle = 180 - (90 + 360 * (1 - 0.5) / 20)
  )

label_nl <-
  country_rank %>%
  filter(country == "Netherlands") %>%
  mutate(
    hjust = medals_per_capita + exp(10),
    angle = 180 - (90 + 360 * (3 - 0.5) / 20)
  )

label_max <-
  country_rank %>%
  slice_min(rank_by_total, n = 10) %>%
  filter(country != "Netherlands") %>%
  mutate(
    rank_id = dense_rank(-rank_mpc),
    hjust = medals_per_capita + exp(10),
    angle = 80 + (360 / 20) * -rank_id
  )

label_min <-
  country_rank %>%
  slice_min(rank_mpc, n = 10) %>%
  filter(country != "Netherlands") %>%
  mutate(
    rank_id = dense_rank(-rank_mpc),
    rank_id = if_else(rank_id == 1, 1, rank_id + 1),
    hjust = medals_per_capita + exp(10),
    angle = 80 + (360 / 20) * -rank_id
  )


country_rank %>%
  slice_min(rank_by_total, n = 10) %>%
  mutate(group = factor("Total medals top 10")) %>%
  bind_rows(country_rank %>%
    slice_max(medals_per_capita, n = 10) %>%
    mutate(group = "Medal per capita top 10")) %>%
  bind_rows(country_rank %>%
    filter(country == "Denmark") %>%
    mutate(group = "Denmark")) %>%
  group_by(group) %>%
  mutate(country = fct_reorder(country, medals_per_capita)) %>%
  ggplot(aes(country, medals_per_capita, fill = group)) +
  geom_col(alpha = .7, width = 0.8, position = "dodge") +
  geom_richtext(
    inherit.aes = FALSE, data = label_dk, alpha = .7,
    aes(country, medals_per_capita, label = name_rank, angle = angle, hjust = -.1),
    fontface = "bold", size = 2.5,
    color = "red"
  ) +
  geom_richtext(
    inherit.aes = FALSE, data = label_max, alpha = .7,
    aes(country, medals_per_capita, label = name_rank, angle = angle, hjust = 1.2),
    fontface = "bold", size = 2.5, show.legend = FALSE, color = "grey25"
  ) +
  geom_richtext(
    inherit.aes = FALSE, data = label_nl, alpha = .7,
    aes(country, medals_per_capita, label = name_rank, angle = angle, hjust = -.1),
    fontface = "bold", size = 2.5,
    color = "grey25"
  ) +
  geom_richtext(
    inherit.aes = FALSE, data = label_min, alpha = .7,
    aes(country, medals_per_capita, label = name_rank, angle = angle, hjust = -.1),
    fontface = "bold", size = 2.5, color = "gray25"
  ) +
  geom_label(aes(10, 0, label = "Medals\nper capita"),
    color = "grey40",
    fontface = "bold",
    fill = "white",
    label.size = NA,
    size = 5
  ) +
  geom_label(aes(19, 8e3, label = "Total medals\ntop 10 countries"),
    fontface = "bold",
    color = "lightblue",
    fill = "white"
  ) +
  geom_label(aes(12.5, 1e3, label = "Medals per capita\ntop 10 countries"),
    fontface = "bold",
    color = "lightgreen",
    fill = "white"
  ) +
  coord_polar() +
  scale_fill_manual(
    values = c("#E22800", "lightgreen", "skyblue"),
  ) +
  scale_y_continuous(
    trans = "log10",
    limits = c(.1, 9600),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))
  ) +
  labs(
    fill = NULL,
    title = NULL,
    y = "Medals per 100 M capita",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    legend.position = "none",
    plot.title = element_text(hjust = .5, size = 20, face = "bold", color = "grey25"),
    axis.title.y = element_text(color = "grey25", hjust = .7)
  )
```
